A robust and memory-efficient transition state search method for complex energy landscapes

### 2022-09-01 manuscript.tex
Halim's changes, consistent hyphons, removed blue

% Temporary
\usepackage{xcolor}
\usepackage{soul}
\soulregister\cite7
\soulregister\cref7
\soulregister\num7
\soulregister\multicolumn7
\renewcommand{\hl}[1]{#1} % Uncomment to remove mark-up

@@

\subsection{Comparison with chain-of-states methods}
\hl{%
In contrast to BITSS, chain-of-states methods do not typically find transition states directly.
Instead, they are designed to find the full pathway (or an approximation thereof), and a secondary method can then be used to refine to the transition state.
As we will demonstrate, this strategy is successful for simple, linear pathways, but faces two key challenges when the pathway is highly non-linear.
Firstly, for such complex pathways, a large number of states are required to sufficiently approximate the minimum energy pathway.
The second is that choosing a suitable initial interpolation can be problematic to achieve.
BITSS is advantageous in both these regards, as only two states are evolved, regardless of the pathway complexity, and no initial interpolation is required.
For these advantages to be of practical use, BITSS must demonstrate at least a comparable speed of convergence to the transition states, relative to these hybrid chain-of-states methods, which we now investigate.

Here we compare the speeds of convergence of BITSS to two widely used approaches for finding transition states that employ chain-of-states methods: climbing-image nudged elastic band (CINEB) \cite{Henkelman2000a}, and DNEB with hybrid eigenvector following (DNEB-HEVF) \cite{Cerjan1981}.
The core of these methods involves minimising the total energy of a chain of states, connected by elastic springs to keep them equally spaced along the transition pathway.
}{\color{blue}
We fix the two end-points at the minima, so the number of states that are minimised is two fewer than the number of states in the chain.
}\hl{%
CINEB modifies the method by altering the behaviour of the state with the highest energy.
The direction of minimisation on this state is inverted along the pathway direction, effectively converting the saddle point into a local minimum.
Alternatively, DNEB-HEVF involves first minimising the chain of states until a criterion is met, and then performing eigenvector-following from the highest energy state, moving uphill along the smallest eigenvector of the Hessian until it reaches the transition state.
It is also possible to combine eigenvector-following with BITSS, which we also include in the comparison for completeness.
}{\color{blue}
Additional implementation details for these methods are included in \cref{sec:HEVFdetails}.

It is also possible to use the string method with a climbing image \cite{E2007} or eigenvector following \cite{Zimmerman2013}; although, in this case, the results are expected to be similar to the nudged-elastic-band methods.
}\hl{%
We note that our aim in this section is to observe how the BITSS method behaves for different systems, rather than providing a comprehensive comparison of the currently available methods, which has been performed in other works \cite{Koslover2007,Sheppard2008}.
}

@@

\begin{table*}[tb]
  \begin{ruledtabular}
  \begin{tabular}{c cc cccc cccc}
    \tworow{System} & \tworow{BITSS} &
      \tworow{\begin{tabular}[c]{@{}c@{}}BITSS-\\HEVF\end{tabular}} &
        \multicolumn{4}{c}{DNEB-HEVF} & \multicolumn{4}{c}{\hl{CINEB}} \\
    \cmidrule(lr){4-7}\cmidrule(lr){8-11}
             &             &             & 3           & 5           & 10          & 20          & \hl{3  } & \hl{5         } & \hl{10         } & \hl{20         } \\
    \hline                                                                                                                                             
    LJ-7     & 148         & 144         & 135         & 138         & 208         & 361         & \hl{30 } & \hl{153       } & \hl{2000       } & \hl{1692       } \\
    Buckling & \num{12866} & \num{14446} & \num{73694} & \num{7315}  & \num{6242}  & \num{12094} & \hl{---} & \hl{\num{8352}} & \hl{\num{19360}} & \hl{\num{77904}} \\
    Wetting  & \num{12100} & \num{17262} & \num{17840} & \num{17721} & \num{20221} & \num{48286} & \hl{---} & \hl{---       } & \hl{---        } & \hl{---        } \\
  \end{tabular}
  \end{ruledtabular}
  \caption{\label{tab:speed}
    Number of potential gradient calculations required to reach the transition state for the three comparison examples.
    \hl{The climbing-image nudged-elastic-band (CINEB) and DNEB with hybrid eigenvector-following (DNEB-HEVF) methods have been run for different numbers of images.
    Convergence is determined to be when the root-mean-square of the gradient at the estimate for the transition state is less than $10^{-4}$.
    The fields left blank indicate that the method has not converged to the correct transition state.
    }
  }
\end{table*}

@@

\hl{%
The results for the three systems are shown in \cref{tab:speed}.
First, we note that for all three systems, using hybrid eigenvector-following does not significantly improve the speed of BITSS.
Indeed, for buckling and especially wetting, HEVF is detrimental to performance.
Next, it is interesting to compare each method's performance between simple and complex pathways.
In contrast to the wetting transition's highly non-linear pathway, the pathways of the LJ-7 rearrangement and the buckling system can be simply tracked following a gradual variation in the order parameters.
For LJ-7, this is the translation of atoms 5 and 6, and for buckling, this is the radial displacement of the centre of the dimple \cite{Panter2019}.
For the simpler pathways, BITSS is generally slower, but for the complex pathways, BITSS is faster.
Moreover, we see that for the wetting example, CINEB does not converge to the transition state because the estimated tangent vector is highly inaccurate due to the non-linearity of the pathway.

For situations where memory is limited, it is important to minimise the number of images used.
}{\color{blue}
However, efficiently finding the TS is challenging for both CINEB and DNEB-HEVF if too few images are used, as observed for the cylindrical buckling with three images.
BITSS, on the other hand, converges using only two images.
}

@@

{\color{blue}
\section{Implementation details for the DNEB-HEVF and BITSS-HEVF methods}\label{sec:HEVFdetails}
In the hybrid eigenvector-following methods, criteria are used to determine when the initial method has converged sufficiently, at which point the eigenvector following begins.
For BITSS, the criterion is when the average of the two states changes during a BITSS step by less than a tenth of the reduction in the separation given by \cref{eq:diteration}.
As for DNEB, the root-mean-square of the total energy gradient of the chain of states is used with a convergence criterion of $10^{-3}$ for the particle cluster and cylindrical buckling systems, or $10^{-5}$ for the striped wetting.
The eigenvector-following method itself was performed using the program OPTIM \cite{OPTIM}.
}

>>>>

% Temporary
\usepackage{xcolor}
\usepackage{soul}
\soulregister\cite7
\soulregister\cref7
\soulregister\num7
\soulregister\multicolumn7
% \renewcommand{\hl}[1]{#1} % Uncomment to remove mark-up

@@

\subsection{Comparison with chain-of-states methods}
\hl{%
In contrast to BITSS, chain-of-states methods do not typically find transition states directly.
Instead, they are designed to find the full pathway (or an approximation thereof), and a secondary method can then be used to refine to the transition state.
As we will demonstrate, this strategy is successful for simple, linear pathways, but faces two key challenges when the pathway is highly non-linear.
Firstly, for such complex pathways, a large number of states are required to sufficiently approximate the minimum energy pathway.
The second is that choosing a suitable initial interpolation can be problematic to achieve.
BITSS can be advantageous in both these regards, as only two states are evolved, regardless of the pathway complexity, and no initial interpolation is required.

Here we compare the speeds of convergence of BITSS to two widely used approaches for finding transition states that employ chain-of-states methods: climbing image nudged elastic band (CINEB) \cite{Henkelman2000a}, and DNEB with hybrid eigenvector following (DNEB-HEVF) \cite{Cerjan1981}.
The core of these methods involves minimising the total energy of a chain of states, connected by elastic springs to keep them equally spaced along the transition pathway.
We fix the two end-points at the minima, so the number of states that are minimised is two fewer than the number of states in the chain.
CINEB modifies the method by altering the behaviour of the state with the highest energy.
The direction of minimisation on this state is inverted along the pathway direction, effectively converting the saddle point into a local minimum.
Alternatively, DNEB-HEVF involves first minimising the chain of states until a convergence criterion is met, and then performing hybrid eigenvector following from the highest energy state, moving uphill along the smallest eigenvector of the Hessian until it reaches the transition state.
For completeness, we also combine hybrid eigenvector following with BITSS and include the results in the convergence comparison.
Additional implementation details for these methods are included in \cref{sec:HEVFdetails}.

It is also possible to use the string method with a climbing image \cite{E2007} or eigenvector following \cite{Zimmerman2013}; although, in this case, the results are expected to be similar to the nudged elastic band methods.
We note that our aim in this section is to observe how the BITSS method behaves for different systems, rather than providing a comprehensive comparison of the currently available methods, which has been performed in other works \cite{Koslover2007,Sheppard2008}.
}

@@

\begin{table*}[tb]
  \begin{ruledtabular}
  \begin{tabular}{c cc cccc cccc}
    \tworow{System} & \tworow{BITSS} &
      \tworow{\begin{tabular}[c]{@{}c@{}}BITSS-\\HEVF\end{tabular}} &
        \multicolumn{4}{c}{DNEB-HEVF} & \multicolumn{4}{c}{\hl{CINEB}} \\
    \cmidrule(lr){4-7}\cmidrule(lr){8-11}
             &             &             & 3           & 5           & 10          & 20          & \hl{3  } & \hl{5         } & \hl{10         } & \hl{20         } \\
    \hline                                                                                                                                             
    LJ-7     & 148         & 144         & 135         & 138         & 208         & 361         & \hl{30 } & \hl{153       } & \hl{2000       } & \hl{1692       } \\
    Buckling & \num{12866} & \num{14446} & \num{73694} & \num{7315}  & \num{6242}  & \num{12094} & \hl{---} & \hl{\num{8352}} & \hl{\num{19360}} & \hl{\num{77904}} \\
    Wetting  & \num{12100} & \num{17262} & \num{17840} & \num{17721} & \num{20221} & \num{48286} & \hl{---} & \hl{---       } & \hl{---        } & \hl{---        } \\
  \end{tabular}
  \end{ruledtabular}
  \caption{\label{tab:speed}
    Number of potential gradient calculations required to reach the transition state for the three comparison examples.
    \hl{The climbing image nudged elastic band (CINEB) and DNEB with hybrid eigenvector following (DNEB-HEVF) methods have been run for different numbers of images.
    Convergence is determined to be when the root-mean-square of the gradient at the estimate for the transition state is less than $10^{-4}$.
    The fields left blank indicate that the method has not converged to the correct transition state.
    }
  }
\end{table*}

@@

\hl{%
The results for the three systems are shown in \cref{tab:speed}.
First, we note that for all three systems, using hybrid eigenvector following does not significantly improve the speed of BITSS.
Indeed, for buckling and especially wetting, HEVF is detrimental to performance.
Next, it is interesting to compare each method's performance between simple and complex pathways.
In contrast to the wetting transition's highly non-linear pathway, the pathways of the LJ-7 rearrangement and the buckling system can be simply tracked following a gradual variation in the order parameters.
For LJ-7, this is the translation of atoms 5 and 6, and for buckling, this is the radial displacement of the centre of the dimple \cite{Panter2019}.
For the simpler pathways, BITSS is generally slower, but for the complex pathways, BITSS is faster.
Moreover, we see that for the wetting example, CINEB does not converge to the transition state because the estimated tangent vector is highly inaccurate due to the non-linearity of the pathway.

For situations where memory is limited, it is important to minimise the number of images used.
However, efficiently finding the TS is challenging for both CINEB and DNEB-HEVF if too few images are used, as observed for the cylindrical buckling with three images.
BITSS, on the other hand, converges using only two images.
}

@@

\hl{%
\section{Implementation details for the CINEB, DNEB-HEVF, and BITSS-HEVF methods}\label{sec:HEVFdetails}
The spring constants used to connect the states in the CINEB and DNEB methods are system dependent.
They are chosen such that they keep the states equidistant without overwhelming the gradients arising from the potential energy landscapes under consideration.
We employ $10^{-1}$ for the Lennard-Jones particle cluster, $10^{-2}$ for the cylindrical buckling example, and $10^{-6}$ for the striped wetting system.

When hybrid eigenvector following is combined with BITSS or DNEB, we need to set out criteria to determine when the double-ended method has sufficiently converged, at which point the hybrid eigenvector following method begins.
For BITSS, the criterion is when the average of the two states changes by less than a tenth of the reduction in the separation given by \cref{eq:diteration} during a BITSS step.
For DNEB, the root-mean-square of the total energy gradient of the chain of states is used with a convergence criterion of $10^{-3}$ for the particle cluster and cylindrical buckling systems, and $10^{-5}$ for the striped wetting.
We employ the hybrid eigenvector following method implemented in the program OPTIM \cite{OPTIM}.
}
